Exploring how changes in cause of death have narrowed the

### 2017-02-10 Manuscript/Manuscript_COD_decomp.Rmd
more changes

**Background**

The difference in life expectancy between black and whites in the US has decreased over time. States have fared differently in their progress towards eliminating the life expectancy gap, with states such as New York undergoing greater reductions to the gap compared to states such as California and Texas. The objective of this study is to compare how changes to cause-specific mortality between 1969 and 2013 have resulted in relatively narrow life expectancy gaps in some states and  wide gaps in other states.

**Methods and Findings**
Using data from the National Vital Statistics System, we extracted the number of deaths and population sizes according to year (1969-2013), state, age, sex, race (black or white), and six cause of death categories (Cancer, cardiovascular disease, communicable disease, non-communicable disease, injuries, and all other causes). We excluded 10 states with the smallest black populations, but included the District of Columbia. A bayesian autoregressive model was used to model smoothed mortality rates over time within each strata, and a truncated Poisson model was used to imputed death counts that were suppressed due to small cell sizes. We calculated life expectancy for blacks and whites within each strata using traditional life table methods, and decomposed the difference in life expectancy by cause of death using Arriaga's method.

@@

**Conclusions**
[Add conclusion]

@@

For each era, we calculated the degree that the life expectancy gap has narrowed or widened over the era and decomposed this *change* to the *difference* in the life expectnacy between blacks and whites according to cause of death. To do so, we simply subtracted the contribution (in years) that each cause of death category had made to the difference in life expectancy at the beginning of the era from that made at the beginning of the era. An example of the calculation is found in the eAppendix.

@@

males.era1 <- ggplotly(
  ggplot(subset(cod.df.males, era == "1969-1983"), aes(y = state.gdiff.num.1, x = new.start.1)) +
    geom_rect(aes(xmin = new.start.1, ymin = state.gdiff.num.1 - 0.45, 
                  ymax = state.gdiff.num.1 + 0.45, 
                  xmax = new.finish.1, fill = COD), 
              color = "white") +
#    facet_grid(.~ Census_Region) +
    geom_segment(data = subset(cod.df.males, COD == "Cancers" & era == "1969-1983"), aes(x = second.gap, xend = second.gap, y = state.gdiff.num.1 - 0.5, yend = state.gdiff.num.1 + 0.5)) + 
        geom_segment(data = subset(cod.df.males, COD == "Cancers" & era == "1969-1983"), aes(x = first.gap, xend = first.gap, y = state.gdiff.num.1 - 0.5, yend = state.gdiff.num.1 + 0.5), lty = 3) + 
   # geom_segment(data = bw_dat.males,
  #               aes(x = second.gap, xend= second.gap, y = state.gdiff.num.1 - 0.5, yend = state.gdiff.num.1 + 0.5)) +
  #  geom_segment(data = bw_dat.males,
  #               aes(x = first.gap, xend= first.gap, y = state.gdiff.num.1 - 0.5, yend = state.gdiff.num.1 + 0.5), lty = 3) +
    theme_minimal() + scale_y_continuous(breaks = 1:y.num.breaks, labels = y.labels)) %>%
  layout(xaxis = list(title = "Life expectancy gap (years): 1969-1983"), yaxis = list(title = NA, autorange = "reversed"))

@@

males.era2 <- ggplotly(
  ggplot(subset(cod.df.males, era == "1983-1993"), aes(y = state.gdiff.num.2, x = new.start.2)) +
    geom_rect(aes(xmin = new.start.2, ymin = state.gdiff.num.2 - 0.45, 
                  ymax = state.gdiff.num.2 + 0.45, 
                  xmax = new.finish.2, fill = COD), 
              color = "white") +
    geom_segment(data = bw_dat.males, 
                 aes(x = third.gap, xend= third.gap, y = state.gdiff.num.2 - 0.5, yend = state.gdiff.num.2 + 0.5)) +
    geom_segment(data = bw_dat.males, 
                 aes(x = second.gap, xend= second.gap, y = state.gdiff.num.2 - 0.5, yend = state.gdiff.num.2 + 0.5), lty = 3) +
    theme_minimal() + scale_y_continuous(breaks = 1:y.num.breaks.2, labels = y.labels.2)) %>% 
  layout(xaxis = list(title = "Life expectancy gap (years): 1969-1983"), yaxis = list(title = NA, autorange = "reversed"))

@@

males.era3 <- ggplotly(
  ggplot(subset(cod.df.males, era == "1993-2013"), aes(y = state.gdiff.num.3, x = new.start.3)) +
    geom_rect(aes(xmin = new.start.3, ymin = state.gdiff.num.3 - 0.45, 
                  ymax = state.gdiff.num.3 + 0.45, 
                  xmax = new.finish.3, fill = COD), 
              color = "white") +
    geom_segment(data = bw_dat.males, 
                 aes(x = fourth.gap, xend= fourth.gap, y = state.gdiff.num.3 - 0.5, yend = state.gdiff.num.3 + 0.5)) +
    geom_segment(data = bw_dat.males, 
                 aes(x = third.gap, xend= third.gap, y = state.gdiff.num.3 - 0.5, yend = state.gdiff.num.3 + 0.5), lty = 3) +
    theme_minimal() + scale_y_continuous(breaks = 1:y.num.breaks.3, labels = y.labels.3)) %>% 
  layout(xaxis = list(title = "Life expectancy gap (years): 1993-2013"), yaxis = list(title = NA, autorange = "reversed"))

@@

**S1 Text: Decomposing the change to the difference in life expectancy over time by cause of death: an example**
To create.

>>>>

**Background:**
The difference in life expectancy between black and whites in the US has decreased over time. States have fared differently in their progress towards eliminating the life expectancy gap, with states such as New York undergoing greater reductions to the gap compared to states such as California and Texas. The objective of this study is to compare how changes to cause-specific mortality between 1969 and 2013 have resulted in relatively narrow life expectancy gaps in some states and  wide gaps in other states.

**Methods and Findings:**
Using data from the National Vital Statistics System, we extracted the number of deaths and population sizes according to year (1969-2013), state, age, sex, race (black or white), and six cause of death categories (Cancer, cardiovascular disease, communicable disease, non-communicable disease, injuries, and all other causes). We excluded 10 states with the smallest black populations, but included the District of Columbia. A bayesian autoregressive model was used to model smoothed mortality rates over time within each strata, and a truncated Poisson model was used to imputed death counts that were suppressed due to small cell sizes. We calculated life expectancy for blacks and whites within each strata using traditional life table methods, and decomposed the difference in life expectancy by cause of death using Arriaga's method.

@@

**Conclusions:**
[Add conclusion]

@@

For each era, we calculated the degree that the life expectancy gap has narrowed or widened over the era and decomposed this *change* to the *difference* in the life expectancy between blacks and whites according to cause of death. To do so, we simply subtracted the contribution (in years) that each cause of death category had made to the difference in life expectancy at the beginning of the era from that made at the beginning of the era. An example of the calculation is found in **S1 Text**.

@@

males.era1 <- ggplotly(
  ggplot(subset(cod.df.males, era == "1969-1983"), aes(y = state.gdiff.num.1, x = new.start.1)) +
    geom_rect(aes(xmin = new.start.1, ymin = state.gdiff.num.1 - 0.45, 
                  ymax = state.gdiff.num.1 + 0.45, 
                  xmax = new.finish.1, fill = COD), 
              color = "white") +
    geom_segment(data = subset(cod.df.males, COD == "Cancers" & era == "1969-1983"), aes(x = second.gap, xend = second.gap, y = state.gdiff.num.1 - 0.5, yend = state.gdiff.num.1 + 0.5)) + 
        geom_segment(data = subset(cod.df.males, COD == "Cancers" & era == "1969-1983"), aes(x = first.gap, xend = first.gap, y = state.gdiff.num.1 - 0.5, yend = state.gdiff.num.1 + 0.5), lty = 3) + 
    theme_minimal() + 
    scale_x_continuous(breaks = seq(2, 16, by = 2), labels = seq(2, 16, by = 2)) +
    scale_y_continuous(breaks = 1:y.num.breaks, labels = y.labels)) %>%
  layout(xaxis = list(title = "Life expectancy gap (years): 1969-1983"), yaxis = list(title = NA, autorange = "reversed"))

@@

males.era2 <- ggplotly(
  ggplot(subset(cod.df.males, era == "1983-1993"), aes(y = state.gdiff.num.2, x = new.start.2)) +
    geom_rect(aes(xmin = new.start.2, ymin = state.gdiff.num.2 - 0.45, 
                  ymax = state.gdiff.num.2 + 0.45, 
                  xmax = new.finish.2, fill = COD), 
              color = "white") +
    geom_segment(data = bw_dat.males, 
                 aes(x = third.gap, xend= third.gap, y = state.gdiff.num.2 - 0.5, yend = state.gdiff.num.2 + 0.5)) +
    geom_segment(data = bw_dat.males, 
                 aes(x = second.gap, xend= second.gap, y = state.gdiff.num.2 - 0.5, yend = state.gdiff.num.2 + 0.5), lty = 3) +
    theme_minimal() + 
    scale_x_continuous(breaks = seq(2, 16, by = 2), labels = seq(2, 16, by = 2)) +
    scale_y_continuous(breaks = 1:y.num.breaks.2, labels = y.labels.2), minor_breaks = NULL) %>% 
  layout(xaxis = list(title = "Life expectancy gap (years): 1969-1983"), yaxis = list(title = NA, autorange = "reversed"))

@@

males.era3 <- ggplotly(
  ggplot(subset(cod.df.males, era == "1993-2013"), aes(y = state.gdiff.num.3, x = new.start.3)) +
    geom_rect(aes(xmin = new.start.3, ymin = state.gdiff.num.3 - 0.45, 
                  ymax = state.gdiff.num.3 + 0.45, 
                  xmax = new.finish.3, fill = COD), 
              color = "white") +
    geom_segment(data = bw_dat.males, 
                 aes(x = fourth.gap, xend= fourth.gap, y = state.gdiff.num.3 - 0.5, yend = state.gdiff.num.3 + 0.5)) +
    geom_segment(data = bw_dat.males, 
                 aes(x = third.gap, xend= third.gap, y = state.gdiff.num.3 - 0.5, yend = state.gdiff.num.3 + 0.5), lty = 3) +
    theme_minimal() + 
    scale_x_continuous(breaks = seq(0, 16, by = 2), labels = seq(0, 16, by = 2)) +
    scale_y_continuous(breaks = 1:y.num.breaks.3, labels = y.labels.3)) %>% 
  layout(xaxis = list(title = "Life expectancy gap (years): 1993-2013"), yaxis = list(title = NA, autorange = "reversed"))

@@

**S1 Text: Decomposing the change to the difference in life expectancy over time by cause of death: an example**

To create.
1. First show the data for NY in 1993 and 2013 for black and white males (smoothed deaths, population, for each age strata).

2. Create life tables for blacks and whites in the two years (4 tables)

3. Calculate the LE gap for each year (two numbers)

4. Decompose by age the BW gap within year (two decomps)

5. Decompose by age and COD the BW gap within year (two decomps)

6. Marginal COD decomp within year (two decomps)

7. Subtract the COD decomp from year 1 from year 2. (one change in decomp)
